\textbf{Generalized additive models to analyze non-linear trends in biomedical longitudinal data using R: Beyond repeated measures ANOVA and Linear Mixed Models}

### 2022-01-19 Manuscripts/Manuscript_by_chapters-SIM_Revisions_final/appendices/Appendix_A.tex
Updated, made changes

%theme colors for the code chunks (originally from latex-solarized on GitHub)
%https://github.com/jez/latex-solarized
\usepackage{xcolor}
\definecolor{sbase03}{HTML}{002B36}
\definecolor{sbase02}{HTML}{073642}
\definecolor{sbase01}{HTML}{586E75}
\definecolor{sbase00}{HTML}{657B83}
\definecolor{sbase0}{HTML}{839496}
\definecolor{sbase1}{HTML}{93A1A1}
\definecolor{sbase2}{HTML}{EEE8D5}
\definecolor{sbase3}{HTML}{FDF6E3}
\definecolor{syellow}{HTML}{B58900}
\definecolor{sorange}{HTML}{CB4B16}
\definecolor{sred}{HTML}{DC322F}
\definecolor{smagenta}{HTML}{D33682}
\definecolor{sviolet}{HTML}{6C71C4}
\definecolor{sblue}{HTML}{268BD2}
\definecolor{scyan}{HTML}{2AA198}
\definecolor{sgreen}{HTML}{859900}
%command to set parameter(s) in package listings
\lstset{
    % How/what to match
    sensitive=true,
    % Border (above and below)
    frame=lines,
    % Extra margin on line (align with paragraph)
    xleftmargin=\parindent,
    % Put extra space under caption
    belowcaptionskip=1\baselineskip,
    % Colors
    backgroundcolor=\color{sbase3},
    basicstyle=\color{sbase00}\ttfamily,
    keywordstyle=\color{scyan},
    commentstyle=\color{sbase1},
    stringstyle=\color{sblue},
    numberstyle=\color{sviolet},
    identifierstyle=\color{sbase00},
    % Break long lines into multiple lines?
    breaklines=true,
    % Show a character for spaces?
    showstringspaces=false,
    tabsize=2
}

@@

#alpha for ribbon in the smooth plots
al <- 0.8

@@

\begin{lstlisting}[language=R]
dat<-tibble(StO2  = c(4, 27, 3, 2, 0.5, 7,4, 50, 45, 56),
            Day   = rep(c(0, 2, 5, 7, 10), times = 2),
            Group = as.factor(rep(c("Control", "Treatment"), each = 5)))
\end{lstlisting}

@@

\begin{lstlisting}[language=R]
gam_00<-gam(StO2_sim ~ s(Day, k = 5),
            method   ='REML',
            data     = dat_sim)
\end{lstlisting}

@@

\begin{lstlisting}[language=R]
gam_01<-gam(StO2_sim ~ s(Day, by=Group, k = 5),
            method ='REML',
            data   = dat_sim)

@@

\begin{lstlisting}[language=R]
#GAM for StO2

gam_02 <- gam(StO2_sim ~ Group+s(Day, by = Group, k = 5),
            method ='REML',
            data   = dat_sim)

@@

\begin{lstlisting}[language=R]
######IMPORTANT: uncommenting the following two lines
    #removes the group means from the comparison######
\end{lstlisting}

@@

\begin{lstlisting}[language=R]
source(here::here("Manuscripts", "Manuscript_by_chapters-SIM_Revisions_final", 
                  "scripts","pointwise_comparisons.R"))
source(here::here("Manuscripts", "Manuscript_by_chapters-SIM_Revisions_final",
                  "scripts", "difference_smooths.R"))
source(here::here("Manuscripts", "Manuscript_by_chapters-SIM_Revisions_final", 
                  "scripts", "pairwise_limits.R"))
#compute difference between smooths and calculate confidence interval: complete data
diff_complete <- difference_smooths(gam_02, smooth = "s(Day)", 
                                    newdata = newdat, 
                                    unconditional = TRUE,
                                    frequentist = FALSE,
                                    n = 100, partial_match = TRUE, 
                                    nrep=10000)
my_list <- pairwise_limits(diff_complete)
\end{lstlisting}

@@

\begin{lstlisting}[language=R]
rib_col <- '#8D7D82'     #color for ribbon for confidence interval
control_rib <- '#875F79' #color for ribbon for control region
treat_rib <- '#A7D89E'   #color for ribbon treatment region

c1 <- ggplot() +
  geom_line(data = diff_complete, aes(x = Day, y = diff), 
            size = 1, alpha = 0.5) +
  annotate("rect", xmin = my_list$init1, xmax = my_list$final1,
           ymin = -Inf, ymax = Inf, fill = control_rib,
           alpha = 0.5) + 
  annotate("text", x = 1.5, y = -18, label = "Control>Treatment",
           size = 6, angle = 90) +
  annotate("rect", xmin = my_list$init2, xmax = my_list$final2,
           ymin = -Inf, ymax = Inf, fill = treat_rib,
           alpha = 0.5) +
  annotate("text", x = 6, y = -18, label = "Treatment>Control",
           size = 6, angle = 90) +
  geom_ribbon(data = diff_complete, 
              aes(x = Day, ymin = lower_s, ymax = upper_s),
              alpha = 0.5, fill = rib_col, inherit.aes = FALSE) +
  geom_hline(yintercept = 0, lty = 2, color = "red") +
  scale_x_continuous(breaks = c(0, 2, 5, 7, 10)) +
  labs(y = "Difference\n(Complete observations)") +
  theme_classic()+
  theme(axis.text = element_text(size = 22))
\end{lstlisting}

>>>>

%theme colors for the code chunks (originally from latex-solarized on GitHub)
%https://github.com/jez/latex-solarized
\usepackage{xcolor}
\definecolor{sbase03}{HTML}{002B36}
\definecolor{sbase02}{HTML}{073642}
\definecolor{sbase01}{HTML}{586E75}
\definecolor{sbase00}{HTML}{657B83}
\definecolor{sbase0}{HTML}{839496}
\definecolor{sbase1}{HTML}{93A1A1}
\definecolor{sbase2}{HTML}{EEE8D5}
\definecolor{sbase3}{HTML}{FDF6E3}
\definecolor{syellow}{HTML}{B58900}
\definecolor{sorange}{HTML}{CB4B16}
\definecolor{sred}{HTML}{DC322F}
\definecolor{smagenta}{HTML}{D33682}
\definecolor{sviolet}{HTML}{6C71C4}
\definecolor{sblue}{HTML}{268BD2}
\definecolor{scyan}{HTML}{2AA198}
\definecolor{sgreen}{HTML}{859900}
%command to set parameter(s) in package listings
\lstset{
    language=R,
    upquote=true,
    % How/what to match
    sensitive=true,
    % Border (above and below)
    frame=lines,
    % Extra margin on line (align with paragraph)
    xleftmargin=\parindent,
    % Put extra space under caption
    belowcaptionskip=1\baselineskip,
    % Colors
    backgroundcolor=\color{sbase3},
    basicstyle=\color{sbase00}\ttfamily,
    keywordstyle=\color{scyan},
    commentstyle=\color{sbase1},
    stringstyle=\color{sblue},
    numberstyle=\color{sviolet},
    identifierstyle=\color{sbase00},
    % Break long lines into multiple lines?
    breaklines=true,
    % Show a character for spaces?
    showstringspaces=false,
    tabsize=2
}

@@

# alpha for ribbon in the smooth plots
al <- 0.8

@@

\begin{lstlisting}[language=R]
dat <- tibble(StO2 = c(4, 27, 3, 2, 0.5, 7, 4, 50, 45, 56), Day = rep(c(0, 2, 5,
    7, 10), times = 2), Group = as.factor(rep(c("Control", "Treatment"), each = 5)))
\end{lstlisting}

@@

\begin{lstlisting}[language=R]
gam_00 <- gam(StO2_sim ~ s(Day, k = 5), method = "REML", data = dat_sim)
\end{lstlisting}

@@

\begin{lstlisting}[language=R]
gam_01 <- gam(StO2_sim ~ s(Day, by = Group, k = 5), method = "REML", data = dat_sim)

@@

\begin{lstlisting}[language=R]
# GAM for StO2

gam_02 <- gam(StO2_sim ~ Group + s(Day, by = Group, k = 5), method = "REML", data = dat_sim)

@@

\begin{lstlisting}[language=R]
###### IMPORTANT: uncommenting the following two lines removes the group means
###### from the comparison######
\end{lstlisting}

@@

\begin{lstlisting}[language=R]
source(here::here("Manuscripts", "Manuscript_by_chapters-SIM_Revisions_final", "scripts",
    "pointwise_comparisons.R"))
source(here::here("Manuscripts", "Manuscript_by_chapters-SIM_Revisions_final", "scripts",
    "difference_smooths.R"))
source(here::here("Manuscripts", "Manuscript_by_chapters-SIM_Revisions_final", "scripts",
    "pairwise_limits.R"))
# compute difference between smooths and calculate confidence interval:
# complete data
diff_complete <- difference_smooths(gam_02, smooth = "s(Day)", newdata = newdat,
    unconditional = TRUE, frequentist = FALSE, n = 100, partial_match = TRUE, nrep = 10000)
my_list <- pairwise_limits(diff_complete)
\end{lstlisting}

@@

\begin{lstlisting}[language=R]
rib_col <- "#8D7D82"  #color for ribbon for confidence interval
control_rib <- "#875F79"  #color for ribbon for control region
treat_rib <- "#A7D89E"  #color for ribbon treatment region

c1 <- ggplot() + geom_line(data = diff_complete, aes(x = Day, y = diff), size = 1,
    alpha = 0.5) + annotate("rect", xmin = my_list$init1, xmax = my_list$final1,
    ymin = -Inf, ymax = Inf, fill = control_rib, alpha = 0.5) + annotate("text",
    x = 1.5, y = -18, label = "Control>Treatment", size = 6, angle = 90) + annotate("rect",
    xmin = my_list$init2, xmax = my_list$final2, ymin = -Inf, ymax = Inf, fill = treat_rib,
    alpha = 0.5) + annotate("text", x = 6, y = -18, label = "Treatment>Control",
    size = 6, angle = 90) + geom_ribbon(data = diff_complete, aes(x = Day, ymin = lower_s,
    ymax = upper_s), alpha = 0.5, fill = rib_col, inherit.aes = FALSE) + geom_hline(yintercept = 0,
    lty = 2, color = "red") + scale_x_continuous(breaks = c(0, 2, 5, 7, 10)) + labs(y = "Difference\n(Complete observations)") +
    theme_classic() + theme(axis.text = element_text(size = 22))
\end{lstlisting}

### 2022-01-19 Manuscripts/Manuscript_by_chapters-SIM_Revisions_final/appendices/Appendix_B.tex
Updated, made changes

#alpha for ribbon in the smooth plots
al<-     0.8

@@

\begin{lstlisting}[language=R]
## This function plots the rm-ANOVA and LMEM for the data simulated in example.R
plot_example <- function(sim_dat) {
  txt <- 20
  p1 <- sim_dat$dat %>%
    ggplot(aes(x = time, y = y, group = treatment, color = treatment)) +
    geom_point(show.legend = FALSE) +
    labs(y = 'response')+
    geom_line(aes(x = time, y = mu, color = treatment), 
              show.legend = FALSE) +
    theme_classic() +
    theme(plot.title = element_text(size = txt, face = "bold"),
          text=element_text(size = txt)) +
    thm1

  #plot the simulated data with trajectories per each subject
  p2 <- sim_dat$dat %>%
    ggplot(aes(x = time, y = y, group = subject, color = treatment)) +
    geom_line(aes(size = "Subjects"), show.legend = FALSE) +
    # facet_wrap(~ treatment) +
    geom_line(aes(x = time, y = mu, color = treatment, 
                  size = "Simulated Truth"),
              lty = 1, show.legend = FALSE) +
    labs(y = 'response') +
    scale_size_manual(name = "Type", 
                      values = c("Subjects" = 0.5, "Simulated Truth" = 3)) +
    theme_classic() +
    theme(plot.title = element_text(size = txt, face = "bold"),
          text = element_text(size = txt)) +
    thm1

  #plot the errors
  p3 <- sim_dat$dat %>%
    ggplot(aes(x = time, y = errors, group = subject, color = treatment)) +
    geom_line(show.legend = FALSE) +
    labs(y = 'errors') +
    theme_classic() +
    theme(plot.title = element_text(size = txt, face = "bold"),
          text = element_text(size = txt))+
    thm1

  #plot the model predictions for rm-ANOVA
  p4 <- ggplot(sim_dat$dat, aes(x = time, y = y, color = treatment)) +
    geom_point(show.legend = FALSE) +
    labs(y = 'response')+
    geom_line(aes(y = predict(sim_dat$fit_anova), 
                  group = subject, 
                  size = "Subjects"),
              show.legend = FALSE) +
    geom_line(data = sim_dat$pred_dat,
              aes(y = predict(sim_dat$fit_anova,
                              level = 0,
                              newdata = sim_dat$pred_dat),
                  size = "Population"),
              show.legend=FALSE) +
    guides(color = guide_legend(override.aes = list(size = 2))) +
    scale_size_manual(name = "Predictions",
                      values=c("Subjects" = 0.5, "Population" = 3)) +
    theme_classic() +
    theme(plot.title = element_text(size = txt, face = "bold"),
          text = element_text(size = txt)) +
    thm1

  #plot the LMEM predictions
  p5 <- ggplot(sim_dat$dat, aes(x = time, y = y, color = treatment)) +
    geom_point()+
    labs(y = 'response')+
    geom_line(aes(y = predict(sim_dat$fit_lme),
                  group = subject, size = "Subjects")) +
    geom_line(data = sim_dat$pred_dat,
              aes(y = predict(sim_dat$fit_lme, 
                              level = 0,
                              newdata = sim_dat$pred_dat),
                  size = "Population")) +
    guides(color = guide_legend(override.aes = list(size = 2)))+
    scale_size_manual(name = "Predictions",
                      values=c("Subjects" = 0.5, "Population" = 3)) +
    theme_classic() +
    theme(plot.title = element_text(size = txt, face = "bold"),
          text = element_text(size=txt))+
    thm1

  if(option=='simple'){
    return((p1 + p4 + p5) + plot_layout(nrow = 1) + 
             plot_annotation(tag_levels = 'A'))
  }
  else {
    return((p1 + p3 + p2 + p4 + p5) + plot_layout(nrow = 1) + 
             plot_annotation(tag_levels = 'A'))
  }
}
\end{lstlisting}

@@

A1 <- plot_example(example(fun_type = "linear", error_type = "correlated"),
                 option = 'composite')

B1 <- plot_example(example(fun_type = "linear", error_type = "independent"),
                 option = 'composite')

C1 <- plot_example(example(fun_type = "quadratic", error_type = "correlated"), 
                 option = 'composite')

D1 <- plot_example(example(fun_type = "quadratic", error_type = "independent"), 
                 option = 'composite') 
\end{lstlisting}

@@

\begin{lstlisting}[language=R]
#data from Vishwanath et. al.
dat <- tibble(StO2 = c(4, 27, 3, 2, 0.5, 7, 4, 50, 45, 56),
              Day = rep(c(0, 2, 5, 7, 10), times = 2),
              Group = as.factor(rep(c("Control", "Treatment"), each = 5)))
\end{lstlisting}

@@

\begin{lstlisting}[language=R]
#This function simulates data for the tumor data using default parameters of 10 observations per time point,and Standard deviation (sd) of 5%.
#Because physiologically StO2 cannot go below 0%, data is
#generated with a cutoff value of 0.0001 (the "StO2_sim")

simulate_data <- function(dat, n = 10, sd = 5) {
  dat_sim <- dat %>%
    slice(rep(1:n(), each = n)) %>%
    group_by(Group, Day) %>%
    mutate(
      StO2_sim = pmax(rnorm(n, StO2, sd), 0.0001),
      subject = rep(1:10),
      subject = factor(paste(subject, Group, sep = "-"))
    ) %>%
    ungroup()

@@

\begin{lstlisting}[language=R]
#the same model used for the full dataset
mod_m1 <- gam(StO2_sim ~ Group+s(Day, by = Group, k = 5), 
              data  = dat_missing)
#appraise the model
appraise(mod_m1)

@@

f6 <- ggplot(ci, aes(x = Day, y = est, group = smooth)) +
  geom_line(lwd = 1) +
  geom_ribbon(data = ci, 
              aes(ymin = lower, ymax = upper, x = Day, 
                  group = smooth, fill = Group),
              inherit.aes = FALSE, alpha = 0.7, show.legend = FALSE) +
  geom_ribbon(data = si,
              aes(ymin = lower, ymax = upper, x = Day, 
                  group = smooth, fill =Group),
              inherit.aes = FALSE, alpha = 0.4, show.legend = TRUE) +
  geom_line(data = si, aes(Day, upper, color = Group), 
            size = 0.8, alpha = 0.7) +
  geom_point(data = dat_missing, aes(x = Day, y = StO2_sim, color=Group), 
             size = 1.5, alpha = 0.6, inherit.aes = FALSE) +
  labs(y = expression(atop(StO[2], '(incomplete observations)'))) +
  scale_x_continuous(breaks = c(0, 2, 5, 7, 10)) +
  theme_classic() +
  theme(axis.text=element_text(size = 22)) +
  thm1
\end{lstlisting}

@@

\begin{lstlisting}[language=R]
#this function calculates the pointwise (by calling pointwise_comparisons.R) CI and the simultaneous CI
# for a pairwise comparison between two smooths
difference_smooths <- function(model, smooth, n = 100, ci_level = 0.95, 
                               newdata = NULL, partial_match = TRUE,
                               unconditional = FALSE, frequentist = FALSE,
                               nrep = 10000, include_means = TRUE, ...) {
  if (missing(smooth)) {
    stop("Must specify a smooth to difference via 'smooth'.")
  }

  # smooths in model
  S <- gratia::smooths(model) # vector of smooth labels - "s(x)"
  # select smooths
  select <-
    gratia:::check_user_select_smooths(smooths = S, select = smooth,
                                       partial_match = partial_match)

@@

  out <- pmap(pairs, difference_pointwise, smooth = smooth, 
              by_var = by_var, smooth_var = smooth_var, data = sm_data,
              Xp = Xp, V = V, coefs = coefs, nrep = nrep)
  out <- bind_rows(out)
  crit <- qnorm((1 - ci_level) / 2, lower.tail = FALSE)

@@

\begin{lstlisting}[language=R]
#function to obtain values for the shading regions of the pairwise comparison between the smooths

pairwise_limits <- function(dataframe) {
  #extract values where the lower limit of the ribbon is greater than zero
  #this is the region where the control group effect is greater

  v1 <- dataframe %>%
    filter(lower_s > 0) %>%
    select(Day)
  #get day  initial value
  init1 = v1$Day[[1]]
  #get day final value
  final1 = v1$Day[[nrow(v1)]]
  #extract values where the value of the upper limit of the ribbon is lower than zero
  #this corresponds to the region where the treatment group effect is greater
  v2 <- dataframe %>%
    filter(upper_s < 0) %>%
    select(Day)

  init2 = v2$Day[[1]]
  final2 = v2$Day[[nrow(v2)]]
  #store values
  my_list <- list(init1 = init1,
                  final1 = final1,
                  init2 = init2,
                  final2 = final2)
  return(my_list)

@@

\begin{lstlisting}[language=R]
#Function that uses the source code of `gam.check` to obtain the estimates without the plots. The source can be checked by typing `gam.check` in the console.
gam_diagnostics <- function (b, old.style = FALSE, 
                           type = c("deviance", "pearson", "response"), 
                           k.sample = 5000, k.rep = 200, rep = 0,
                           level = 0.9, rl.col = 2, rep.col = "gray80", 
                           ...) {
  type <- match.arg(type)
  resid <- residuals(b, type = type)
  linpred <- if (is.matrix(b$linear.predictors) && !is.matrix(resid))
    napredict(b$na.action, b$linear.predictors[, 1])
  else napredict(b$na.action, b$linear.predictors)

>>>>

# alpha for ribbon in the smooth plots
al <- 0.8

@@

\begin{lstlisting}[language=R]
## This function plots the rm-ANOVA and LMEM for the data simulated in
## example.R
plot_example <- function(sim_dat) {
    txt <- 20
    p1 <- sim_dat$dat %>%
        ggplot(aes(x = time, y = y, group = treatment, color = treatment)) + geom_point(show.legend = FALSE) +
        labs(y = "response") + geom_line(aes(x = time, y = mu, color = treatment),
        show.legend = FALSE) + theme_classic() + theme(plot.title = element_text(size = txt,
        face = "bold"), text = element_text(size = txt)) + thm1

    # plot the simulated data with trajectories per each subject
    p2 <- sim_dat$dat %>%
        ggplot(aes(x = time, y = y, group = subject, color = treatment)) + geom_line(aes(size = "Subjects"),
        show.legend = FALSE) + # facet_wrap(~ treatment) + show.legend = FALSE)
        show.legend = FALSE) + # facet_wrap(~ treatment) + + # facet_wrap(~
        show.legend = FALSE) + # facet_wrap(~ treatment) + treatment) +
    geom_line(aes(x = time, y = mu, color = treatment, size = "Simulated Truth"),
        lty = 1, show.legend = FALSE) + labs(y = "response") + scale_size_manual(name = "Type",
        values = c(Subjects = 0.5, `Simulated Truth` = 3)) + theme_classic() + theme(plot.title = element_text(size = txt,
        face = "bold"), text = element_text(size = txt)) + thm1

    # plot the errors
    p3 <- sim_dat$dat %>%
        ggplot(aes(x = time, y = errors, group = subject, color = treatment)) + geom_line(show.legend = FALSE) +
        labs(y = "errors") + theme_classic() + theme(plot.title = element_text(size = txt,
        face = "bold"), text = element_text(size = txt)) + thm1

    # plot the model predictions for rm-ANOVA
    p4 <- ggplot(sim_dat$dat, aes(x = time, y = y, color = treatment)) + geom_point(show.legend = FALSE) +
        labs(y = "response") + geom_line(aes(y = predict(sim_dat$fit_anova), group = subject,
        size = "Subjects"), show.legend = FALSE) + geom_line(data = sim_dat$pred_dat,
        aes(y = predict(sim_dat$fit_anova, level = 0, newdata = sim_dat$pred_dat),
            size = "Population"), show.legend = FALSE) + guides(color = guide_legend(override.aes = list(size = 2))) +
        scale_size_manual(name = "Predictions", values = c(Subjects = 0.5, Population = 3)) +
        theme_classic() + theme(plot.title = element_text(size = txt, face = "bold"),
        text = element_text(size = txt)) + thm1

    # plot the LMEM predictions
    p5 <- ggplot(sim_dat$dat, aes(x = time, y = y, color = treatment)) + geom_point() +
        labs(y = "response") + geom_line(aes(y = predict(sim_dat$fit_lme), group = subject,
        size = "Subjects")) + geom_line(data = sim_dat$pred_dat, aes(y = predict(sim_dat$fit_lme,
        level = 0, newdata = sim_dat$pred_dat), size = "Population")) + guides(color = guide_legend(override.aes = list(size = 2))) +
        scale_size_manual(name = "Predictions", values = c(Subjects = 0.5, Population = 3)) +
        theme_classic() + theme(plot.title = element_text(size = txt, face = "bold"),
        text = element_text(size = txt)) + thm1

    if (option == "simple") {
        return((p1 + p4 + p5) + plot_layout(nrow = 1) + plot_annotation(tag_levels = "A"))
    } else {
        return((p1 + p3 + p2 + p4 + p5) + plot_layout(nrow = 1) + plot_annotation(tag_levels = "A"))
    }
}
\end{lstlisting}

@@

A1 <- plot_example(example(fun_type = "linear", error_type = "correlated"), option = "composite")

B1 <- plot_example(example(fun_type = "linear", error_type = "independent"), option = "composite")

C1 <- plot_example(example(fun_type = "quadratic", error_type = "correlated"), option = "composite")

D1 <- plot_example(example(fun_type = "quadratic", error_type = "independent"), option = "composite")
\end{lstlisting}

@@

\begin{lstlisting}[language=R]
# data from Vishwanath et. al.
dat <- tibble(StO2 = c(4, 27, 3, 2, 0.5, 7, 4, 50, 45, 56), Day = rep(c(0, 2, 5,
    7, 10), times = 2), Group = as.factor(rep(c("Control", "Treatment"), each = 5)))
\end{lstlisting}

@@

\begin{lstlisting}[language=R]
# This function simulates data for the tumor data using default parameters of
# 10 observations per time point,and Standard deviation (sd) of 5%. Because
# physiologically StO2 cannot go below 0%, data is generated with a cutoff
# value of 0.0001 (the 'StO2_sim')

simulate_data <- function(dat, n = 10, sd = 5) {
    dat_sim <- dat %>%
        slice(rep(1:n(), each = n)) %>%
        group_by(Group, Day) %>%
        mutate(StO2_sim = pmax(rnorm(n, StO2, sd), 1e-04), subject = rep(1:10), subject = factor(paste(subject,
            Group, sep = "-"))) %>%
        ungroup()

@@

\begin{lstlisting}[language=R]
# the same model used for the full dataset
mod_m1 <- gam(StO2_sim ~ Group + s(Day, by = Group, k = 5), data = dat_missing)
# appraise the model
appraise(mod_m1)

@@

f6 <- ggplot(ci, aes(x = Day, y = est, group = smooth)) + geom_line(lwd = 1) + geom_ribbon(data = ci,
    aes(ymin = lower, ymax = upper, x = Day, group = smooth, fill = Group), inherit.aes = FALSE,
    alpha = 0.7, show.legend = FALSE) + geom_ribbon(data = si, aes(ymin = lower,
    ymax = upper, x = Day, group = smooth, fill = Group), inherit.aes = FALSE, alpha = 0.4,
    show.legend = TRUE) + geom_line(data = si, aes(Day, upper, color = Group), size = 0.8,
    alpha = 0.7) + geom_point(data = dat_missing, aes(x = Day, y = StO2_sim, color = Group),
    size = 1.5, alpha = 0.6, inherit.aes = FALSE) + labs(y = expression(atop(StO[2],
    "(incomplete observations)"))) + scale_x_continuous(breaks = c(0, 2, 5, 7, 10)) +
    theme_classic() + theme(axis.text = element_text(size = 22)) + thm1
\end{lstlisting}

@@

\begin{lstlisting}[language=R]
# this function calculates the pointwise (by calling pointwise_comparisons.R)
# CI and the simultaneous CI for a pairwise comparison between two smooths
difference_smooths <- function(model, smooth, n = 100, ci_level = 0.95, newdata = NULL,
    partial_match = TRUE, unconditional = FALSE, frequentist = FALSE, nrep = 10000,
    include_means = TRUE, ...) {
    if (missing(smooth)) {
        stop("Must specify a smooth to difference via 'smooth'.")
    }

    # smooths in model
    S <- gratia::smooths(model)  # vector of smooth labels - 's(x)'
    # select smooths
    select <- gratia:::check_user_select_smooths(smooths = S, select = smooth, partial_match = partial_match)

@@

    out <- pmap(pairs, difference_pointwise, smooth = smooth, by_var = by_var, smooth_var = smooth_var,
        data = sm_data, Xp = Xp, V = V, coefs = coefs, nrep = nrep)
    out <- bind_rows(out)
    crit <- qnorm((1 - ci_level)/2, lower.tail = FALSE)

@@

\begin{lstlisting}[language=R]
# function to obtain values for the shading regions of the pairwise comparison
# between the smooths

pairwise_limits <- function(dataframe) {
    # extract values where the lower limit of the ribbon is greater than zero
    # this is the region where the control group effect is greater

    v1 <- dataframe %>%
        filter(lower_s > 0) %>%
        select(Day)
    # get day initial value
    init1 = v1$Day[[1]]
    # get day final value
    final1 = v1$Day[[nrow(v1)]]
    # extract values where the value of the upper limit of the ribbon is lower
    # than zero this corresponds to the region where the treatment group effect
    # is greater
    v2 <- dataframe %>%
        filter(upper_s < 0) %>%
        select(Day)

    init2 = v2$Day[[1]]
    final2 = v2$Day[[nrow(v2)]]
    # store values
    my_list <- list(init1 = init1, final1 = final1, init2 = init2, final2 = final2)
    return(my_list)

@@

\begin{lstlisting}[language=R]
# Function that uses the source code of `gam.check` to obtain the estimates
# without the plots. The source can be checked by typing `gam.check` in the
# console.
gam_diagnostics <- function(b, old.style = FALSE, type = c("deviance", "pearson",
    "response"), k.sample = 5000, k.rep = 200, rep = 0, level = 0.9, rl.col = 2,
    rep.col = "gray80", ...) {
    type <- match.arg(type)
    resid <- residuals(b, type = type)
    linpred <- if (is.matrix(b$linear.predictors) && !is.matrix(resid))
        napredict(b$na.action, b$linear.predictors[, 1]) else napredict(b$na.action, b$linear.predictors)

### 2022-01-19 Manuscripts/Manuscript_by_chapters-SIM_Revisions_final/sections/03-GAM Theory.Rmd
Updated, made changes

To get the weighed basis functions, each basis (from Figure \@ref(fig:basis-plot)A) is multiplied by the corresponding coefficients in Figure \@ref(fig:basis-plot)B, thereby increasing or decreasing the original basis functions. Figure \@ref(fig:basis-plot)C shows the resulting weighed basis functions. Note that the magnitude of the weighting for the first basis function has resulted in a decrease of its overall contribution to the smoother term (because the coefficient for that basis function is negative and its magnitude is less than one). On the other hand, the third basis function has roughly doubled its contribution to the smooth term. Finally, the weighted basis functions are added at each timepoint to produce the smooth term. The resulting smooth term for the effect of _time_ is shown in Figure \@ref(fig:basis-plot)D (brown line), along the simulated values which appear as points.

(ref:basis-plot-caption) Basis functions for a single smoother for time. **A**: Basis functions for a single smoother for time for the simulated data of Group 1 from Figure 2. **B**: Matrix of basis function weights. Each basis function is multiplied by a coefficient which can be positive or negative. The coefficient determines the overall effect of each basis in the final smoother. **C**: Weighed basis functions. Each of the four basis functions (and the intercept) of panel A has been weighed by the corresponding coefficient shown in Panel B. Note the corresponding increase (or decrease) in magnitude of each weighted basis function. **D**: Smoother for time and original data points. The smoother (line) is the result of the sum of each weighted basis function at each time point, with simulated values for Group 1 shown as points.

@@

Besides the across the function CI, "simultaneous" or "whole function" CIs can also be computed, which contain the _whole function_ with a specified probability [@wood2017]. Suppose we chose a nominal value (say, 95%) and compute a simultaneous CI; if we obtain 100 repeated samples and compute a simultaneous CI in each case, we would expect that the true function lies completely within the computed simultaneous CI in 95 of those repeated samples. Briefly, to obtain a simultaneous CI we use simulation from the empirical Bayesian posterior distribution to obtain the maximum absolute standardized deviation of the model estimates, which is used to correct the coverage of the across the function CI [@ruppert2003].

>>>>

To get the weighted basis functions, each basis (from Figure \@ref(fig:basis-plot)A) is multiplied by the corresponding coefficients in Figure \@ref(fig:basis-plot)B, thereby increasing or decreasing the original basis functions. Figure \@ref(fig:basis-plot)C shows the resulting weighted basis functions. Note that the magnitude of the weighting for the first basis function has resulted in a decrease of its overall contribution to the smoother term (because the coefficient for that basis function is negative and its magnitude is less than one). On the other hand, the third basis function has roughly doubled its contribution to the smooth term. Finally, the weighted basis functions are added at each timepoint to produce the smooth term. The resulting smooth term for the effect of _time_ is shown in Figure \@ref(fig:basis-plot)D (brown line), along the simulated values which appear as points.

(ref:basis-plot-caption) Basis functions for a single smoother for time. **A**: Basis functions for a single smoother for time for the simulated data of Group 1 from Figure 2. **B**: Matrix of basis function weights. Each basis function is multiplied by a coefficient which can be positive or negative. The coefficient determines the overall effect of each basis in the final smoother. **C**: Weighted basis functions. Each of the four basis functions (and the intercept) of panel A has been weighted by the corresponding coefficient shown in Panel B. Note the corresponding increase (or decrease) in magnitude of each weighted basis function. **D**: Smoother for time and original data points. The smoother (line) is the result of the sum of each weighted basis function at each time point, with simulated values for Group 1 shown as points.

@@

Besides the across the function CI, "simultaneous" or "whole function" CIs can also be computed, which contain the _whole function_ with a specified probability [@wood2017]. Suppose we chose a nominal value (say, 95%) and compute a simultaneous CI; if we obtain 100 repeated samples and compute a simultaneous CI in each case, we would expect that the true function lies completely within the computed simultaneous CI in 95 of those repeated samples. Briefly, to obtain a simultaneous CI we simulate 10,000 draws from the empirical Bayesian posterior distribution of the fitted smooths. Then, we obtain the maximum absolute standardized deviation of the differences in smooth estimates which is used to correct the coverage of the across the function CI [@ruppert2003] in a similar fashion to how _q-values_ correct _p-values_ to control false positive discovery rates [@storey2003].

### 2022-01-19 Manuscripts/Manuscript_by_chapters-SIM_Revisions_final/sections/04-Longitudinal analysis with GAMs.Rmd
Updated, made changes

At the core of a biomedical longitudinal study lies the question of a significant difference between the effect of two or more treatments in different groups. Whereas in rm-ANOVA a _post-hoc_ analysis is required to answer such question by calculating some _p-values_ after multiple comparisons, GAMs can use a different approach to estimate significance. In essence, the idea behind the estimation of significance in GAMs across different treatment groups is that the difference between the separate smoothers per group (such as in `gam_02`) can be computed, followed by the estimation of an empirical Bayesian simultaneous CI around this difference.

@@

The absence of a _p-value_ in this context might seem odd, but the pairwise difference in smooths can be conceptualized in the following manner: Different time trends in each group are an indication of an effect by the treatment as in Figure \@ref(fig:sim-smooth-plot)A, where the chemotherapy causes $\mbox{StO}_2$ to increase over time.

With this expectation of different trends in each group, computing the difference between the trends will identify if the observed difference is significant. A difference between groups with similar trends is unlikely to be distinguishable from zero, which would indicate that the treatment is not causing a change in the response in one of the groups (assuming the other group is a Control or Reference group). Therefore, the computation of both the difference between the smooths and the corresponding simultaneous empirical Bayesian CI around this difference is able to provide an estimation of when and by how much there is a significant difference between the different groups. Additionally, the correction provided by the simultaneous empirical Bayesian CI makes the estimation robust as we know that on average, the simultaneous CI will contain the _whole function_ at a nominal value (say, 95%).

>>>>

At the core of a biomedical longitudinal study lies the question of a significant difference between the effect of two or more treatments in different groups. In linear models (such as rm-ANOVA), if there is a significant _p-value_ after a _post-hoc_ analysis we then can make inference about the effect size using the slope or the intercept from the model. In GAMs however, there is no single _p-value_ to determine the significance of an effect as in linear models. Therefore, the coefficients of GAMs do not provide a simple interpretation as in the linear model case, but the changes in slope at specific timepoints can be used to determine the instantaneous effect size. In essence, the idea behind the estimation of significance in GAMs across different treatment groups is that the difference between the separate smoothers per group (such as in `gam_02`) can be computed pairwise, followed by the estimation of an empirical Bayesian simultaneous CI around this difference.

@@

The pairwise difference in smooths can be conceptualized in the following manner: Different time trends in each group are an indication of an effect by the treatment as in Figure \@ref(fig:sim-smooth-plot)A, where the chemotherapy causes $\mbox{StO}_2$ to increase over time. With this expectation of different trends in each group, computing the difference between the trends will identify if the observed difference is significant. A difference between groups with similar trends is unlikely to be distinguishable from zero, which would indicate that the treatment is not causing a change in the response in one of the groups (assuming the other group is a Control or Reference group). Therefore, the computation of both the difference between the smooths and the corresponding simultaneous empirical Bayesian CI around this difference is able to provide an estimation of when and by how much there is a significant difference between the different groups. Additionally, the correction provided by the simultaneous empirical Bayesian CI makes the estimation robust as we know that on average, the simultaneous CI will contain the _whole function_ at a nominal value (say, 95%).
